Towards a wide-coverage grammar for Swedish using GF

### 2011-12-28 doc/report/report.tex
report: elisabets corrections

\abstract{
This thesis describes the work to achieve a covering Swedish grammar implemented
in GF -- to model natural language by functional programming. 
The idea has been to combine existing language technology resources with
new techniques, and the aim to obtain a system for analysing written text.
To reach this goal, problems of computational as well as linguistic nature had to be solved.
\\The thesis explains the methods and technologies used and further inspects the possibilities
of reaching our long-term goal; to accomplish a parser for unrestricted Swedish, by
combining our rule-based grammar with statistical methods.
\\Our contribution is a wide-coverage GF lexicon, a translation of a Swedish treebank
to GF notation and we also give an implementation Swedish grammar. The grammar is based
on the multilingual abstract syntax given in GF resource library, and now
extends it to cover constructions specific to Swedish.
We further give an example of the advantage of using dependent types when
describing grammar and syntax, in this case reflexive pronouns.

@@

\newpage
\pagestyle{empty}
\section*{Acknowledgments}
\vspace{5mm}
Many people have helped me during this work and made this thesis possible.
I would first of all like to thank
Center of Language Technology, that has funded the project.\\

@@

I am also grateful to Krasimir Angelov, Markus Forsberg, Peter Ljunglöf, Lars Borin
and many others who have contributed with ideas and inspiration, and for their
interest in this work.\\

Finally, I would also like to thank my friends and family. Special thanks to Dan Rosén
for all his support, advice and patience and, most importantly, for being such a good friend. %for having listened to 
%my monologues about the daily progress. 
% Ramona, Aarne, Elisabet, Markus, Peter?, Lars, Dan, Krasimir etc
% Funding!

% check references! lncs is good maybe. lnai
\newpage
%\pagestyle{empty}
%\cleardoublepage
\tableofcontents
%
\addtocontents{toc}{\protect\thispagestyle{empty}}
%\pagestyle{empty}
%\cleardoublepage
\newpage
\setcounter{page}{1} %reset the page counter
\chapter{Introduction}
% more intro to gf. dependently typed...high-level formalism
Grammatical Framework \cite{gfbok} is a dependently typed grammar formalism,
developed for multilinguality. It is based on Martin-Löf type theory which allows
reasoning within the programming language.
GF has so far been successfully used for 
controlled languages \cite{cnl}, but recent experiments have showed
that it is also possible to use the framework for parsing free language \cite{patent}.
% >> although the success rate is going steeply down when using uncontrolled nl. Better with annotated.
% --> sentence about connection GF - grammar based parser.
This project focuses on implementing a wide-coverage Swedish grammar, which also
describes a parser accepting all sentences that can be expressed by the grammar.

Parsing, that is the task of automatically identifying morphological and
syntactical structures, is subject to increasing interest, especially
considering the steadily growing amount of data available online. 
So far there is no freely available grammar-driven parser that gives a deep
analysis for Swedish. The fact that %the existing rule-based Swedish
a parser or grammar is not freely available does not only restrict its
usage but also its possibilities of being developed and our
goal is
to create a open source Swedish grammar. % from which we derive a parser
%accepting all sentences described by the given rules.
As a freely available resource, it can be continuously enhanced and
made use of in other projects.\\
To build the grammar from scratch would be not only time consuming but
would also mean that already existing resources would reimplemented.
In order to not reinvent the wheel we proceed from a combination of well-tested
resources.
We start from GF resource grammar and thereby get
%The usage of GF allows us to start from
a well-defined system for describing grammar, as well as a strong connection to
more than 20 other languages implemented in the framework. Further, we use
the extensive lexicon Saldo and the treebank Talbanken.
%>>The grammar defines a parser, which accepts all constructions defined by the grammar.

\section{Aims}
%>> enhance gf...
The purpose of this project has been to prepare the GF grammar for free parsing.
This has meant to develop earlier techniques to fit for Swedish, create methods
for achieving and keeping a large-scale lexicon and to adapt the existing
resource grammar to fit for more language specific constrictions.
Three sub-goals has been identified:
\begin{itemize}
\item Enhancing the Swedish GF grammar
\item Importing the lexicon Saldo
\item Create a translation between Talbanken and GF
\end{itemize}
%We hence needed to extend the Swedish grammar to cover
%the language specific constructions. Further a large lexicon was
%needed, as well as a method for evaluation.
%During this project, we have worked on and explored those things and
%while identifying problems and future work.

\newpage % until it looks nicer
\section{Outline}
The thesis is divided into 6 chapters. We start by giving some background
information to areas relevant to the project. %Chapter \ref{sec:background} 
Grammatical Framework is presented in section \ref{sec:gf} while a more profound
description of the Swedish resource grammar is given in section \ref{sec:swegf}.
Section \ref{sec:swedish} gives an introduction Swedish, and
brief presentations of Saldo and Talbanken are found in section \ref{sec:saldo} and
\ref{sec:talbanken} respectively.
A summary of related work can be found in section \ref{sec:related}.

@@

Finally, the conclusion and evaluation are
presented in chapter \ref{sec:results} together with a list of future work .

@@

\chapter{Background}  
\label{sec:background}
The work described in this is thesis is part of a bigger project which aims at
using GF for parsing free Swedish. 
In previous work\footnote{??}, a start was made to extending the Swedish GF grammar
and a tool for lexical acquisition was developed.
We now construct a bigger
and more expressive grammar as well as a large scale GF lexicon.
As all GF grammars, this one defines a parser, and we develop it by getting
examples, ideas and test material from the treebank Talbanken.
The project is hence heavily depending on three resources, which will be described
in this section.

@@

GF is a strongly typed functional programming language, inspired by
%programming languages like 
ML \cite{ml} and Haskell \cite{haskell}. It is also a logical framework,
and the built-in functionality for logic and reasoning 
is inspired by \textlambda Prolog \cite{prolog} and 
by Agda \cite{agda}, with which GF also shares its type checking algorithm.
The first version of the framework was implemented by Xerox Research Center
in Grenoble and is now mainly developed in Gothenburg. One of the biggest
achievements is a library covering the 
basic morphological and syntactic structures of more than
20 languages (see section \ref{sec:resources}).

%> dependent and implicit types : agda. (read gf book and krasses book)
A grammar written in GF can be used for both parsing and generation.
The parsing algorithm is incremental and has polynomial time and space
complexity \cite{gfMech}. 
The GF package also provides various tools for working with and using the grammars:
a compiler, an interpreter and a runtime system.
The grammars can be compiled into portable grammar format (PGF) \cite{pgf},
supported by %There are libraries for using PGF files with
Java, Java script, C and Haskell libraries. 
%>> Haskell only complete, also or C\# and C.  %may also be used in Android applications.
The interpreter, the GF shell, allows
the user to test grammars by commands for parsing, visualization of parse trees,
random generation, word alignment, morphological quizzes etc.
An interactive GF editor and the GF shell can be tried out online 
\footnote{http://www.grammaticalframework.org/demos/gfse/}.
\begin{figure}[h]
\begin{verbatim} 
> parse "jag ser katten" | visualize_parse -format=pdf -view=evince
> parse "jag ser katten" | visualize_tree  -format=pdf -view=evince
\end{verbatim}
\caption{Example of how the GF shell can be used. The first command will parse the given
sentence and visualize the resulting parse tree. The second will visualize the 
abstract syntax tree. The trees will be showed in the pgf-format by the program evince}
\label{fig:shellvp}
\end{figure}
\begin{figure}[h]
\centering
\subfloat[GF abstract tree]{\label{pic:gfAtree}\includegraphics[width=100mm]{gfTree.pdf}}
\subfloat[Swedish parse tree]{\label{pic:gfStree}\includegraphics[width=30mm]{gfSTree.png}}
\caption{Parse tree and abstract tree for the sentence ``Jag ser katten".}
\label{fig:gftree1}
\end{figure}

% nice: test vt -format=pdf -view=evince -nofun
\begin{wrapfigure}{l}{0.22\textwidth}
\includegraphics[width=20mm]{gfETree.png}
\caption{English parse tree}
\label{pic:gfEtree}
\end{wrapfigure}
The visualization of the trees is done with graphviz\footnote{http://www.graphviz.org/}.
Figure \ref{fig:gftree1} shows the results from running the commands in figure \ref{fig:shellvp}
%A GF parse tree has two visualized representations, the \textit{parse tree} and
%the \textit{abstract tree}.
The \textit{parse tree} (\ref{pic:gfStree}) shows which types
that are assigned to the words and phrases. 
The \textit{abstract tree} shows which functions GF uses
to parse a sentence.
% do not know about abstract/concrete yet: and is consequently the same for all concrete syntaxes.
The reader should keep in mind that the visualized parse trees is not a
complete representation. It does not reproduce all information from the abstract tree.
%If the end node of a  branch has no corresponding word, it is excluded from
%the visualized parse tree. This is the case for the 
For example, the definiteness and number of the noun \emph{`katten'}, which is shown as
\verb-DetQuant DefArt NumSg- in the abstract tree.
%and the information \verb-DetQuant DefArt NumSg- is not shown in figure \ref{pic:gfStree}.
In the corresponding English parse tree, figure \ref{pic:gfEtree},
the noun is explicitly quantified by the article \emph{``the"},
and the determiner is therefore shows in the parse tree.\\

@@

%>>the scope is controlled language, but possible ambiguities (discuss this in future
%>>work, saldo, conclusion). Ambiguities: from multilinguality, or inherited:
%>>syntactical (man with the telescope), lexical (youSg/Pl, växa).
%>> have ambiguity section. 
\newpage %as long as english picture bråks.
The scope of GF grammars is usually controlled language. By
restricting the coverage, the number of ambiguities
can be limited and we can ensure that the semantics is preserved during
translation. Inherited ambiguities as well as ambiguities arising from multilinguality will
remain, but can be controlled more easily. %Example of domainspecific word?
The use of controlled language thus gives a possibility of good and reliable translation
and GF has successfully been used in project like %Some examples of projects using GF are 
WebAlt \cite{webalt}, a project which aims to develop language independent
material for mathematics problems, and the formal software verification tool
KeY \cite{key}, in which GF is used for translating specification to natural language.
%>> key, talk, hats, bom (see references in Ramonas thesis).
%>> use talk here and move key to other place??
The project talk \cite{talk} used GF to describe dialog grammars and
the framework is further used in the
European project MOLTO \footnote{http://www.molto-project.eu/} for online
translation between up to 15 language.\\

@@

When implementing a English version of the grammar, we encounter another
problem: the verb form depends on the number of the noun. We solve this by letting
\verb|N'| carry
information about which its number and letting \verb|Pred| pass this on to the
verb. Finally, the type of \verb|V| is put to a table, showing the verbs forms for each
number.
\begin{figure}[h!]
\begin{verbatim}              
concrete TestGrammarEng of TestGrammar = {
  lincat S  = Str ;
         V  = {s : Num => Str} ;
         N  = {s : Num => Str} ;
         N' = {s : Str ; num : Num} ;
  lin   
    Pred n v = n.s ++ v.s ! n.num ;
    NPl n = {s = n.s ! Pl ; num = Pl} ;
    NSg n = {s = n.s ! Sg ; num = Sg} ;
    cat_N = {s = table {Sg => "the cat" ;
                        Pl => "the cats"}};
    sleep_V = {s = table {Sg => "sleeps" ; 
                          Pl => "sleep"}};
  param Num = Sg | Pl ;
  }
\end{verbatim}           
\caption{English implementation}
\label{fig:gfTestEng}
\end{figure}

@@

We now have two implementations of the abstract. The resulting GF grammar is able both to
parse a string to a abstract tree and to go in the other direction; to produce
a string of natural language given an abstract tree. This step is called linearization.
Translation is a consequence of this, we can parse a Swedish string and then 
linearize the resulting abstract tree to English.

\subsection{The resource library}
\label{sec:resources}
%>>Best part? Biggest library?
%>>Most general resource, first large grammar, covers fundamental syntactics.
The GF package provides an advantageous resource library \cite{gf-resource}, covering the
fundamental morphology and syntax of more than 20 languages.
There is also a small test lexicon included, containing a few hundred common
words.
%is a very useful part of the framework. It is provided in the 
%, they give a good start for all who want to write an application grammar.
%of a language for each application, all
%this can be found as libraries. So far there are 20 languages in the library, and
The languages all share the same abstract and are valuable
%they constitute  a very useful resource 
when implementing application grammars.

The resource grammars describe how to construct phrases and sentences and how to
decline words. The latter is done by smart paradigms: functions that analyse
some given forms of a word to find out how the inflection table should look.
For example, the declination of many Swedish nouns can be determined by looking
at the singular form only. 
\begin{wrapfigure}{l}{0.4\textwidth}
\begin{tabular}{| l | l |}
\hline
1st declination & 5th declination \\
\hline
flick\textbf{a}    &     hjärt\textbf{a}   \\
flick\textbf{an}    &    hjärt\textbf{at}  \\
flick\textbf{or}    &    hjärt\textbf{an}  \\
flick\textbf{orna}  &    hjärt\textbf{ana} \\
\hline
\end{tabular}
\end{wrapfigure}
This is the case for ``flicka", the correct declination
can be guessed by using the first declination. % and noting that the word ends with an ``a". 
%Like most Swedish nouns ending with an ``a", it follows the first declination.
For others, like "hjärta", also the plural form "hjärtan" is needed.
%, even though it ends with an ``a". 
The worst case is nouns that need four forms, both singular and
plural in definite and indefinite form.
Section \ref{sec:swegf} will give a more thorough description of the Swedish resource grammar.

\subsection{Frontiers of Grammatical Framework}
As an open source-project, GF is constantly being developed and improved. New
languages are added, the compiler is being improved, ways of using it in more 
efficient and easy-going manners are added
and the possibilities to use GF in different environments
increased. There are research on how to make more use of the dependent
types, for reasoning by using ontologies \cite{ramona} or generating natural
language via Montague semantics \cite{montague}.
%Further, experiments on using GF for free parsing has been conducted
%\cite[\textsection 4.1]{gfMech}.
%This is also a future goal for our project and one of the subgoal has been to
%model an important part of Swedish in GF. The starting point was the resource
%grammar to which the result has a strong connection.

@@

\section{Talbanken}
\label{sec:talbanken}
For testing and evaluation of the grammar and lexicon, we needed to be able to
compare them against a reliable source.
Being a freely available, manually annotated, large-scale treebank,
Talbanken \cite{talbanken} was perfect for our purpose.
%It was created in the 1970s at Lund University.
In total it consists of four
parts. Two of them are transcriptions of spoken language, one a collection of
text written by high school students, and one, section P,
consists of professionally written Swedish gathered from newspapers, brochures and textbooks.\\
Talbanken was also used to train the Swedish version of the Maltparser \cite{malt}
and was then redistributed in a more modern version, Talbanken05 \cite{talbanken05}.
%This work was done by Nivre, Hall and Nilsson in 2005.
It is released in Malt\footnote{http://w3.msi.vxu.se/~nivre/research/MaltXML.html} 
and Tiger\footnote{http://www.ims.uni-stuttgart.de/projekte/TIGER/TIGERSearch/doc/html/TigerXML.html}
XML-formats
where the trees had been made deeper and more detailed while still containing
the lexical MAMBA layer. \\
In this project, as well as for the Malt parser, the more than 6000 sentences
in section P has been used.
The treebank has served as an inspiration and an evaluation source throughout the
project. An automatic mapping between its trees and the abstract trees from GF has been
done, which will be explained in section \ref{sec:Mapping}.
%Differences in analyse, do we want a similar? 
%Uses of Talbanken in project, mapping to evaluate and test.  
%Look at output from Maltparser, could also comparing our parse trees to this.

\section{Saldo}
\label{sec:saldo}
A good parser needs a good lexicon. We have used of Saldo \cite{saldo}, a
large electronic lexicon developed and maintained at Gothenburg University. It is
built on Svenskt Associations Lexicon and contains information about more than 
120 000 modern Swedish words.
For each word there is semantic, syntactical and a morphological
information. The user can find compounding analyses, examples of usage in corpora,
and graphs of semantically connected words.\\
\begin{figure}[h!]
\centering
\subfloat[Morphological info for ``katt"]{\label{pic:saldoTab}\includegraphics[width=40mm]{saldotab.png}}
\hspace{20mm}
\subfloat[Graph showing the hyponyms of ``katt"]{\label{pic:saldoWheel}\includegraphics[width=70mm]{saldograph.png}}
\caption{Information in Saldo}
\label{fig:saldo}
\end{figure}

The semantic aspect of Saldo require that words with multiple meanings are
separated into different entries.
%If one word has different meanings, there will be separate entries for each of them. 
The word ``växa" (``grow") for example, has two meanings: \emph{getting
bigger}
\enumsentence{Gräset växte varje dag\\
`The grass grew each day'}
or \emph{being a plant}. 
\enumsentence{Det växte en gammal ek där.\\
`There stood an old oak.'}
``växa" is however declined the same way for both interpretations of the word.
Hence \textit{Saldo's morphological lexicon} only keeps one entry of ``växa".
For the homonyms that do not share the whole inflection tables, different entries are 
kept also in the morphological lexicon.

For our purpose, only this morphological section, provided under LPGL in XML
format, was needed.
The data can be processed and translated to GF format as described in 
section \ref{sec:prog.saldo}.

@@

\subsection*{Post-nominal articles} 
\label{sec:swedishnoun}
Unlike most of the world's languages Swedish express not only the number but also
the definiteness of a noun by suffixes. The endings are decided by the noun's gender,
%which may be 
\textit{neuter} or \textit{non-neuter}.\\
%(or \textit{utrum}) 
\begin{tabular}{l|lll}
&\textbf{Definite}& \textbf{Indefinite}& \textbf{Gender} \\
 \hline
\textbf{Singular} & {en katt} & {katten} & \textit{non-neuter}\\
& \emph{a cat} & \emph{the cat} &\\
\textbf{Plural} & katter & katterna &\\
 &\emph{cats} & \emph{the cats} &\\
 \hline
\textbf{Singular} & {ett hus} & {huset} & \textit{neuter}\\
 & \emph{a house} & \emph{the house} & \\
\textbf{Plural} & hus & husen & \\
& \emph{houses} & \emph{the houses} & \\
\end{tabular}\\

>>>>

\abstract{
This thesis describes the work to achieve a covering Swedish grammar implemented
in GF -- to model natural language by functional programming. 
The idea has been to combine existing language technology resources with
new techniques, and the aim to obtain a system for analysing written text.
To reach this goal, problems of computational as well as linguistic nature had to be solved.
\\The thesis explains the methods and technologies used and further inspects the possibilities
of reaching our long-term goal; to accomplish a parser for unrestricted Swedish, by
combining our rule-based grammar with statistical methods.
\\Our contribution is a wide-coverage GF lexicon, a translation of a Swedish treebank
to GF notation and a Swedish grammar implementation. The grammar is based
on the multilingual abstract syntax given in GF resource library, and now
extends it to cover constructions specific to Swedish.
We further give an example of the advantage of using dependent types when
describing grammar and syntax, in this case reflexive pronouns.

@@

\newpage
\pagestyle{empty}
\section*{Acknowledgments}
\vspace{5mm}
Many people have helped me during this work and made this thesis possible.
I would first of all like to thank
Center of Language Technology, that has funded the project.

@@

I am also grateful to Krasimir Angelov, Markus Forsberg, Peter Ljunglöf, Lars Borin
and many others who have contributed with ideas and inspiration, and for their
interest in this work.

Finally, I would also like to thank my friends and family. Special thanks to Dan % Rosén
for all his support, advice and patience and, most importantly, for being such a good friend. %for having listened to 
%my monologues about the daily progress. 
% Ramona, Aarne, Elisabet, Markus, Peter?, Lars, Dan, Krasimir etc
% Funding!

% check references! lncs is good maybe. lnai
\newpage
%\pagestyle{empty}
%\cleardoublepage
\tableofcontents
%
\addtocontents{toc}{\protect\thispagestyle{empty}}
%\pagestyle{empty}
%\cleardoublepage
\newpage
\setcounter{page}{1} %reset the page counter
\chapter{Introduction}
% more intro to gf. dependently typed...high-level formalism
Grammatical Framework \cite{gfbok} is a dependently typed grammar formalism,
developed for multilinguality. It is based on Martin-Löf type theory which allows
reasoning within the programming language.
GF has so far been successfully used for 
controlled languages \cite{cnl}, but recent experiments have showed
that it is also possible to use the framework for parsing free language \cite{patent}.
% >> although the success rate is going steeply down when using uncontrolled nl. Better with annotated.
% --> sentence about connection GF - grammar based parser.
This project focuses on implementing a wide-coverage Swedish grammar, which
describes a parser accepting all sentences that can be expressed by the grammar.

Parsing, that is the task of automatically identifying morphological and
syntactical structures, is subject to increasing interest, especially
considering the steadily growing amount of data available online. 
So far there is no freely available grammar-driven parser that gives a deep
analysis for Swedish. The fact that %the existing rule-based Swedish
a parser or grammar is not freely available does not only restrict its
usage but also its possibilities of being developed and our
goal is
to create an open source Swedish grammar. % from which we derive a parser
%accepting all sentences described by the given rules.
As a freely available resource, it can be continuously enhanced and
made use of in other projects.\\
To build the grammar from scratch would be not only time consuming but
would also mean that already existing resources would have to be reimplemented.
In order to not reinvent the wheel we proceed from a combination of well-tested
resources.
We start from a GF resource grammar and thereby get
%The usage of GF allows us to start from
a well-defined system for describing the grammar, as well as a strong connection to
more than 20 other languages implemented in the framework. Further, we use
the extensive lexicon Saldo and the treebank Talbanken.
%>>The grammar defines a parser, which accepts all constructions defined by the grammar.

\section{Aims}
%>> enhance gf...
The purpose of this project has been to prepare the GF grammar for parsing of 
unrestricted Swedish.
This has meant to develop earlier techniques to fit for Swedish, create methods
for achieving and keeping a large-scale lexicon and to adapt the existing
resource grammar to fit more language specific constrictions.
Three sub-goals has been identified:
\begin{itemize}
\item Extending the Swedish GF grammar
\item Importing the lexicon Saldo
\item Creating translation between Talbanken and GF
\end{itemize}
%We hence needed to extend the Swedish grammar to cover
%the language specific constructions. Further a large lexicon was
%needed, as well as a method for evaluation.
%During this project, we have worked on and explored those things and
%while identifying problems and future work.

\newpage % until it looks nicer
\section{Outline}
The thesis is divided into 6 chapters. We start by giving some background
information to areas relevant to the project. %Chapter \ref{sec:background} 
Grammatical Framework is presented in section \ref{sec:gf} while a more profound
description of the Swedish resource grammar is given in section \ref{sec:swegf}.
Section \ref{sec:swedish} gives an introduction to the Swedish language, and
brief presentations of Saldo and Talbanken are found in section \ref{sec:saldo} and
\ref{sec:talbanken} respectively.
A summary of related work can be found in section \ref{sec:related}.

@@

Finally, the conclusion and evaluation are
presented in chapter \ref{sec:results} together with some areas of future work.

@@

\chapter{Background}  
\label{sec:background}
The work described in this thesis is part of a bigger project which aims at
using GF for parsing unrestricted Swedish. 
In previous work\footnote{??}, a start was made to extend the Swedish GF grammar
and a tool for lexical acquisition was developed.
We now construct a bigger
and more expressive grammar as well as a large scale GF lexicon.
As all GF grammars, this one defines a parser, and we develop it by getting
examples, ideas and test material from the treebank Talbanken.
The project is hence heavily depending on three resources, which will be described
in this section.

@@

GF is a strongly typed functional programming language, inspired by
%programming languages like 
ML \cite{ml} and Haskell \cite{haskell}. It is also a logical framework,
and the built-in functionality for logic and reasoning 
is inspired by \textlambda Prolog \cite{prolog} and 
by Agda \cite{agda}, with which GF also shares its type checking algorithm.
The first version of the framework was implemented at Xerox Research Center
in Grenoble and is now mainly developed in Gothenburg. One of the biggest
achievements is a library covering the 
basic morphological and syntactic structures of more than
20 languages (see section \ref{sec:resources}).

%> dependent and implicit types : agda. (read gf book and krasses book)
A grammar written in GF can be used for both parsing and generation.
The parsing algorithm is incremental and has polynomial time and space
complexity \cite{gfMech}. 
The GF package also provides various tools for working with and using the grammars:
a compiler, an interpreter and a runtime system.
The grammars can be compiled into portable grammar format (PGF) \cite{pgf},
supported by %There are libraries for using PGF files with
Java, Java script, C and Haskell libraries. 
%>> Haskell only complete, also or C\# and C.  %may also be used in Android applications.
The interpreter, the GF shell, allows
the user to test grammars by commands for parsing, visualization of parse trees,
random generation, word alignment, morphological quizzes etc.
An interactive GF editor and the GF shell can be tried out online 
\footnote{http://www.grammaticalframework.org/demos/gfse/}. See figure \ref{fig:shell},
 The first command will parse the given
sentence and visualize the resulting parse tree. The second will visualize the 
abstract syntax tree. The trees will be showed in the pgf-format by the program evince.
\begin{figure}[h]
\begin{verbatim} 
> parse "jag ser katten" | visualize_parse -format=pdf -view=evince
> parse "jag ser katten" | visualize_tree  -format=pdf -view=evince
\end{verbatim}
\label{fig:shell}
\caption{Example of how the GF shell can be used.}\label{fig:shellvp}
\end{figure}
\begin{figure}[h]
\centering
\subfloat[GF abstract tree]{\label{pic:gfAtree}\includegraphics[width=100mm]{gfTree.pdf}}
\subfloat[Swedish parse tree]{\label{pic:gfStree}\includegraphics[width=30mm]{gfSTree.png}}
\caption{Abstract tree and parse tree for the sentence ``Jag ser katten".}
\label{fig:gftree1}
\end{figure}

% nice: test vt -format=pdf -view=evince -nofun
\begin{wrapfigure}{l}{0.22\textwidth}
\includegraphics[width=20mm]{gfETree.png}
\caption{English parse tree}
\label{pic:gfEtree}
\end{wrapfigure}
The visualization of the trees is done with graphviz\footnote{http://www.graphviz.org/}.
Figure \ref{fig:gftree1} shows the results from running the commands in figure \ref{fig:shellvp}
%A GF parse tree has two visualized representations, the \textit{parse tree} and
%the \textit{abstract tree}.
The \textit{abstract tree} shows which functions GF uses
to parse a sentence.
The \textit{parse tree} (\ref{pic:gfStree}) shows which types
that are assigned to the words and phrases. 
% do not know about abstract/concrete yet: and is consequently the same for all concrete syntaxes.
The reader should keep in mind that the visualized parse trees is not a
complete representation. It does not reproduce all information from the abstract tree.
%If the end node of a  branch has no corresponding word, it is excluded from
%the visualized parse tree. This is the case for the 
For example, the definiteness and number of the noun \emph{`katten'}, which is shown as
\verb-DetQuant DefArt NumSg- in the abstract tree.
%and the information \verb-DetQuant DefArt NumSg- is not shown in figure \ref{pic:gfStree}.
In the corresponding English parse tree, figure \ref{pic:gfEtree},
the noun is explicitly quantified by the article \emph{``the"},
and the determiner is therefore shown in the parse tree.\\

@@

%>>the scope is controlled language, but possible ambiguities (discuss this in future
%>>work, saldo, conclusion). Ambiguities: from multilinguality, or inherited:
%>>syntactical (man with the telescope), lexical (youSg/Pl, växa).
%>> have ambiguity section. 
\newpage %as long as english picture bråks.
The scope of GF grammars is usually controlled language, a limited language used in
a restricted domain. By
restricting the coverage, the number of ambiguities
can be limited and we can ensure that the semantics is preserved during
translation. Inherited ambiguities as well as ambiguities arising from multilinguality will
remain, but can be controlled more easily. %Example of domainspecific word?
The use of controlled language thus gives a possibility of good and reliable translation
and GF has successfully been used in project like %Some examples of projects using GF are 
WebAlt \cite{webalt}, a project which aims to develop language independent
material for mathematics problems. The formal software verification tool
KeY \cite{key}, in which GF is used for translating specification to natural language.
%>> key, talk, hats, bom (see references in Ramonas thesis).
%>> use talk here and move key to other place??
GF can be used for describe dialog grammars see \cite{talk}, and
the framework is further used in the
European project MOLTO \footnote{http://www.molto-project.eu/} for online
translation between up to 15 language.\\

@@

When implementing an English version of the grammar, we encounter another
problem: the verb form depends on the number of the noun. We solve this by letting
\verb|N'| carry
information about its number and letting \verb|Pred| pass this on to the
verb. Finally, the type of \verb|V| is put into a table, showing the verbs forms for each
number.
\begin{figure}[h!]
\begin{verbatim}              
concrete TestGrammarEng of TestGrammar = {
  lincat S  = Str ;
         V  = {s : Num => Str} ;
         N  = {s : Num => Str} ;
         N' = {s : Str ; num : Num} ;
  lin   
    Pred n v = n.s ++ v.s ! n.num ;
    NPl n = {s = n.s ! Pl ; num = Pl} ;
    NSg n = {s = n.s ! Sg ; num = Sg} ;
    cat_N = {s = table {Sg => "the cat" ;
                        Pl => "the cats"}};
    sleep_V = {s = table {Sg => "sleeps" ; 
                          Pl => "sleep"}};
  param Num = Sg | Pl ;
  }
\end{verbatim}           
\caption{English implementation}
\label{fig:gfTestEng}
\end{figure}

@@

We now have two implementations of the abstract, one for Swedish and one for English.
The resulting GF grammar is able both to
parse a string to an abstract tree and to go in the other direction; to produce
a string of natural language given an abstract tree. This step is called linearization.
Translation is a consequence of this, we can parse a Swedish string and then 
linearize the resulting abstract tree to English.

\subsection{The resource library}
\label{sec:resources}
%>>Best part? Biggest library?
%>>Most general resource, first large grammar, covers fundamental syntactics.
The GF package provides an useful resource library \cite{gf-resource}, covering the
fundamental morphology and syntax of more than 20 languages.
There is also a small test lexicon included, containing a few hundred common
words.
%is a very useful part of the framework. It is provided in the 
%, they give a good start for all who want to write an application grammar.
%of a language for each application, all
%this can be found as libraries. So far there are 20 languages in the library, and
The languages all share the same abstract syntax which is  valuable
%they constitute  a very useful resource 
when implementing multilingual application grammars.

The resource grammars describe how to construct phrases and sentences and how to
decline words. The latter is done by smart paradigms: functions that analyse
some given forms of a word to find out how the inflection table should look.
For example, the declination of many Swedish nouns can be determined by looking
at the singular form only. 
\begin{wrapfigure}{l}{0.4\textwidth}
\begin{tabular}{| l | l |}
\hline
1st declination & 5th declination \\
\hline
flick\textbf{a}    &     hjärt\textbf{a}   \\
flick\textbf{an}    &    hjärt\textbf{at}  \\
flick\textbf{or}    &    hjärt\textbf{an}  \\
flick\textbf{orna}  &    hjärt\textbf{ana} \\
\hline
\end{tabular}
\end{wrapfigure}
This is the case for ``flicka", a noun belonging to the first declination. %the correct declination
%can be guessed by using the first declination. % and noting that the word ends with an ``a". 
%Like most Swedish nouns ending with an ``a", it follows the first declination.
For others, like "hjärta", also the plural form "hjärtan" is needed.
%, even though it ends with an ``a". 
The worst case is nouns that need four forms, both singular and
plural in definite and indefinite form.
Section \ref{sec:swegf} will give a more thorough description of the Swedish resource grammar.

\subsection{Frontiers of Grammatical Framework}
As an open source-project, GF is constantly being developed and improved. New
languages are added, the compiler is being improved, ways of using it in more 
efficient and easy-going manners are added
and the possibilities to use GF in different environments
increased. There is research on how to make more use of the dependent
types, for reasoning by using ontologies \cite{ramona} or generating natural
language via Montague semantics \cite{montague}.
%Further, experiments on using GF for free parsing has been conducted
%\cite[\textsection 4.1]{gfMech}.
%This is also a future goal for our project and one of the subgoal has been to
%model an important part of Swedish in GF. The starting point was the resource
%grammar to which the result has a strong connection.

@@

\section{Talbanken}
\label{sec:talbanken}
For testing and evaluation of the grammar and lexicon, we needed to be able to
compare them against a reliable source.
Talbanken \cite{talbanken} was perfect for our purpose,
being a freely available, manually annotated, large-scale treebank.
%It was created in the 1970s at Lund University.
It is analyzed with the MAMBA annotation scheme (Teleman, 1974) and 
consists of four
parts. Two of them are transcriptions of spoken language, one a collection of
text written by high school students, and one, section P,
consists of professionally written Swedish gathered from newspapers, brochures and textbooks.
\\
Talbanken was also used to train the Swedish version of the Malt parser \cite{malt}
and was then redistributed in an updated version,
Talbanken05 \cite{talbanken05}.
%This work was done by Nivre, Hall and Nilsson in 2005.
It is released in Malt\footnote{http://w3.msi.vxu.se/~nivre/research/MaltXML.html} 
and Tiger\footnote{http://www.ims.uni-stuttgart.de/projekte/TIGER/TIGERSearch/doc/html/TigerXML.html}
XML-formats
where the trees had been made deeper and more detailed while still containing
the lexical MAMBA layer. \\
The Malt parser was trained on section P of Talbanken, and these
more than 6000 sentences have been used our project. 
The treebank has served as an inspiration and an evaluation source throughout the
project. An automatic mapping between its trees and the abstract trees from GF has been
done, which will be explained in section \ref{sec:Mapping}.
%Differences in analyse, do we want a similar? 
%Uses of Talbanken in project, mapping to evaluate and test.  
%Look at output from Maltparser, could also comparing our parse trees to this.

\section{Saldo}
\label{sec:saldo}
A good parser needs a good lexicon. We have used of Saldo \cite{saldo}, a
large electronic lexicon developed and maintained at Gothenburg University. It is
built on Svenskt Associationslexikon and contains information about more than 
120 000 modern Swedish words.
For each word there is semantic, syntactical and a morphological
information. The user can find examples of usage in corpora,
graphs of semantically connected words and some suggestions for how to analyse compounds. \\
\begin{figure}[h!]
\centering
\subfloat[Morphological info for ``katt"]{\label{pic:saldoTab}\includegraphics[width=40mm]{saldotab.png}}
\hspace{20mm}
\subfloat[Graph showing the hyponyms of ``katt"]{\label{pic:saldoWheel}\includegraphics[width=70mm]{saldograph.png}}
\caption{Information in Saldo}
\label{fig:saldo}
\end{figure}

The semantic aspect of Saldo requires that words with multiple meanings are
separated into different entries.
%If one word has different meanings, there will be separate entries for each of them. 
The word ``växa" (``grow") for example, has two meanings: \emph{getting
bigger}
\enumsentence{Gräset växte varje dag\\
`The grass grew each day'}
or \emph{being a plant}. 
\enumsentence{Det växte en gammal ek där.\\
`There stood an old oak.'}
``växa" is however declined the same way for both interpretations of the word.
Hence \textit{Saldo's morphological lexicon} only keeps one entry of ``växa".
For the homonyms that do not share the whole inflection tables, different entries are 
kept also in the morphological lexicon.

For our purpose, only the morphological section, provided under LPGL in XML
format, was needed.
The data can be processed and translated to GF format as described in 
section \ref{sec:prog.saldo}.

@@

\subsection*{Post-nominal articles} 
\label{sec:swedishnoun}
Unlike most of the world's languages Swedish express not only the number but also
the definiteness of a noun by suffixes. The endings are decided by the noun's gender,
%which may be 
\textit{neuter} or \textit{non-neuter}.\\
%(or \textit{utrum}) 
\begin{tabular}{l|lll}
                  &\textbf{Indefinite} & \textbf{Definite}& \textbf{Gender} \\
 \hline
\textbf{Singular} & {en katt}        & {katten}           & \textit{non-neuter}\\
                  & \emph{a cat}     & \emph{the cat}     &\\
\textbf{Plural}   & katter           & katterna           &\\
                  &\emph{cats}       & \emph{the cats}    &\\
 \hline
\textbf{Singular} & {ett hus}        & {huset}            & \textit{neuter}\\
                  & \emph{a house}   & \emph{the house}   & \\
\textbf{Plural}   & hus              & husen              & \\
                  & \emph{houses}    & \emph{the houses}  & \\
\end{tabular}\\
